The Ranked Nearest Neighbour Interchange space

### 2019-08-12 paper.tex
some edits in section Geometry

We will identify leaves with their labels and also call them \emph{taxa} (singular: \emph{taxon}) following the systematic biology tradition.
All nodes that are no leafs are called \emph{internal nodes} and edges $(u,v)$ are \emph{internal edge}, if both $u$ and $v$ are internal nodes.
Our definition of ranked tree leads to a natural characterisation of edge lengths within ranked trees:
integer $l$ denotes the \emph{length} of an edge $e=(u,v)$ if $|\rank(v) - \rank(u)| = l$.
Two trees are called \emph{isomorphic} if the underlying rooted phylogenetic trees are isomorphic and the isomorphism preserves ranks of the nodes.

@@

This paper is an advance towards the general goal of understanding the geometry of the $\rnni$ space.
In this section we are investigating the geometry by analysing the structure of shortest paths in $\rnni$.
We will start by presenting an algorithm for computing paths in $\rnni$ (Algorithm~\ref{alg:find_path}) that turns out to be an essential tool for our studies.
Using this algorithm, we are able to prove that there is always a shortest path between two caterpillar trees that only consists of caterpillar trees (Theorem~\ref{thm:caterpillar_convex}) in Section~\ref{section:caterpillar_convex}.
In other words, the set of caterpillar trees is convex in $\rnni$.
This distinguishes $\rnni$ from $\nni$ where this does not hold.
One further result, presented in Section~\ref{section:diameter}, directly following from the proof of Theorem~\ref{thm:caterpillar_convex}, is the diameter of $\rnni$.
Finally, we disprove the so-called Split Theorem, conjectured in \autocite{Gavryushkin2018-ol}, and formulate an alternative version of it, the Cluster Theorem (Section~\ref{section:split_theorem}).

@@

The idea of the algorithm is described in the following.
Each node $v$ of a tree defines a \emph{cluster} $C$, which is the set of taxa descending from the $v$.
We then say that $v$ \emph{induces} the cluster $C$.
Note that the list of all clusters of a tree, if sorted increasingly according to the rank of the corresponding nodes, uniquely defines that tree.
For computing a path from trees $T$ to $R$, we consider clusters $C_1, \ldots, C_{n-1}$ of $R$ where $C_i$ is induced by the node with rank $i$ for $i = 1, \ldots, n-1$.
The so-called \emph{trivial clusters} that are induced by leaves and the root are the same in every tree and do therefore not need to be changed on any path between two trees.
Therefore, the list of clusters $C_1, \ldots, C_{n-1}$, which is ordered increasingly according to the ranks of the corresponding nodes, uniquely defines $R$.
Starting at tree $\hat{T} = T$, $\findpath$ constructs a path where the clusters of $R$ are constructed from bottom to top, using following procedure:
For each $C_i$, find the \emph{most recent common ancestor} $\mrca_{\hat{T}}{C_i}$ (also known as \emph{least common ancestor}) of $C_i$ in current tree $\hat{T}$, that is the internal node with lowest rank that is ancestral to all the taxa in $C_i$.
Update $\hat{T}$ by decreasing the rank of $\mrca_{\hat{T}}{C_i}$ by an $\rnni$ move until it has the same rank $i$ as in $R$. \todo{Do we need to explain why there is always an $\rnni$ move that moves the $\mrca$ down?}
Since we consider the clusters of $R$ such that the ranks of the corresponding internal nodes in $R$ increase, $\mrca_{\hat{T}}{C_i}$ is the parent of cluster $C_i$ in $\hat{T}$ after these $\rnni$ moves.
Thus, repeating this procedure for all clusters of $R$ results in a path between any pair of tress $T, R$, indeed.
Further details, including pseudo-code for $\findpath$, can be found in Section~\ref{section:computation_findpath}.

@@

\begin{lemma}
    $\Delta(\rnni) \leq \frac{(n-1)(n-2)}{2}$, where $\Delta(G)$ denotes the diameter of graph $G$.
\end{lemma}

\begin{proof}
    Let us consider the maximum length of a path given by Algorithm~\ref{alg:find_path}.
    In worst case each of the $n-2$ clusters $C$ induced by internal nodes of $R$ with ranks $i = 1, \dots, n-2$, requires $n-1-i$ $\rnni$ moves to move $\mrca(C)$ down in $\hat{T}$.
	Hence, the maximum length of a path computed by that algorithm for any pair of trees is less or equal to $\sum\limits_{i = 1}^{n-2} i = \frac{(n-2)(n-1)}{2}$.
\end{proof}

Lemma~\ref{lemma:distance_delete_taxon} is central for some of the following proofs using induction.

Central for the following inductive proofs is the comparison of the length of a shortest paths $p$ on trees on $n+1$ taxa with a path on $n$ taxa induced by $p$.
The following lemma bounds the length of these induced paths utilising the rank differences of certain internal nodes between start and end tree of $p$.
Henceforth the notion $\parent(x)$ refers to the internal node that is parent of taxon $x$.

@@

\begin{proof}
    Let $p$ be a shortest path from tree $T$ to $R$ and $p{\big|}_n$ the path resulting from deleting taxon $n+1$ from all trees on $p$.
    This new path $p{\big|}_n$ is then a path from $T$ to $R$.
    Showing that at least $\delta$ $\rnni$ moves are deleted from $p$ to receive $p{\big|}_n$ results in $d(T{\big|}_n,R{\big|}_n) \leq |p{\big|}_n| \leq |p| - \delta = d(T,R) -\delta$ which proves the lemma.
    By the following two observations one can easily see that this claim is true, indeed:
    The rank of an internal node $v$ only changes within an $\rnni$ move if this move is a rank move of $v$ with another node or an $\nni$ move on an edge incident to $v$.
    Furthermore, any $\rnni$ move can change the rank of an internal node by at most one.
    Applying this to $\parent(n+1)$ and knowing that the rank of this node differs by $\delta$ between $T$ and $R$ proves that $d(T{\big|}_n, R{\big|}_n) \leq d(T,R) - \delta$.

@@

Related to Lemma~\ref{lemma:distance_delete_taxon} is Proposition~\ref{proposition:lower_bound_distance} which provides a lower bound for the distance of two trees.
Using the same arguments as in the proof of Lemma~\ref{lemma:distance_delete_taxon}, it easily follows that the rank difference of parents of any taxon in two trees gives such a bound.

@@

If this conjecture is true, we have that $\Delta(\rnni) = rad(\rnni)$.

>>>>

We will identify leaves with their labels and also call them \emph{taxa} (singular: \emph{taxon}) following the systematic biology tradition.
All nodes that are no leafs are called \emph{internal nodes} and an edge $(u,v)$ is an \emph{internal edge}, if both $u$ and $v$ are internal nodes.
Our definition of ranked tree leads to a natural characterisation of edge lengths within ranked trees:
integer $l$ denotes the \emph{length} of an edge $e=(u,v)$ if $|\rank(v) - \rank(u)| = l$.
Two trees are called \emph{isomorphic} if the underlying rooted phylogenetic trees are isomorphic and the isomorphism preserves ranks of the nodes.

@@

This paper is an advance towards the general goal of understanding the geometry of the $\rnni$ space.
In this section we are investigating the geometry by analysing the structure of shortest paths in $\rnni$.
We will start by presenting an algorithm for computing paths in $\rnni$ ($\findpath$) that turns out to be an essential tool for our studies.
Using this algorithm, we are able to prove that there is always a shortest path between two caterpillar trees that only consists of caterpillar trees (Theorem~\ref{thm:caterpillar_convex}) in Section~\ref{section:caterpillar_convex}.
In other words, the set of caterpillar trees is convex in $\rnni$.
This distinguishes $\rnni$ from $\nni$ where this does not hold.
One further result, presented in Section~\ref{section:diameter}, directly following from the proof of Theorem~\ref{thm:caterpillar_convex}, is the diameter of $\rnni$.
Finally, we disprove the so-called Split Theorem, conjectured in \autocite{Gavryushkin2018-ol}, and formulate an alternative version of it, the Cluster Theorem (Section~\ref{section:split_theorem}).

@@

Before introducing the idea of this algorithm, we need further definitions.
Each node $v$ of a tree defines a \emph{cluster} $C$, which is the set of taxa descending from $v$.
We then say that $v$ \emph{induces} $C$.
\emph{Trivial clusters} are clusters induced by leaves and the root of a tree.
Note that the list of all clusters of a tree, if sorted increasingly according to the rank of the corresponding nodes, uniquely defines that tree.
For computing a path from trees $T$ to $R$ we consider clusters $C_1, \ldots, C_{n-1}$ of $R$ where $C_i$ is induced by the node with rank $i$ for $i = 1, \ldots, n-1$.
Trivial clusters do not need to be changed on any path between two trees which is why they are not considered in the algorithm.
This list $C_1, \ldots, C_{n-1}$ of clusters, which is ordered increasingly according to the ranks of their corresponding nodes, uniquely defines $R$.
Starting at tree $\hat{T} = T$, $\findpath$ constructs a path where the clusters of $R$ are constructed from bottom to top, using following procedure:
For each $C_i$, find the \emph{most recent common ancestor} $\mrca_{\hat{T}}(C_i)$ (also known as \emph{least common ancestor}) of $C_i$ in current tree $\hat{T}$, that is the internal node with lowest rank that is ancestral to all the taxa in $C_i$.
Update $\hat{T}$ by decreasing the rank of $\mrca_{\hat{T}}(C_i)$ with an $\rnni$ move until it has the same rank $i$ as in $R$. \todo{Do we need to explain why there is always an $\rnni$ move that moves the $\mrca$ down?}
Since we consider the clusters of $R$ such that the ranks of the corresponding internal nodes in $R$ increase, $\mrca_{\hat{T}}(C_i)$ is the parent of cluster $C_i$ in $\hat{T}$ after these $\rnni$ moves.
Thus, repeating this procedure for all clusters of $R$ results in a path between any pair of tress $T, R$, indeed.
Further details, including pseudo-code for $\findpath$, can be found in Section~\ref{section:computation_findpath}.

@@

\begin{lemma}
    $\Delta(\rnni) \leq \frac{(n-1)(n-2)}{2}$, where $\Delta(G) = \max\limits_{u,v \in V(G)}d(u,v)$ denotes the diameter of graph $G$.
\end{lemma}

\begin{proof}
    Let us consider the maximum length of a path computed by $\findpath$.
    In worst case each of the $n-2$ clusters $C$ induced by internal nodes of $R$ with rank $i = 1, \dots, n-2$, requires $n-1-i$ $\rnni$ operations to move $\mrca(C)$ down in $\hat{T}$.
	Hence, the maximum length of a path computed by that algorithm for any pair of trees is less or equal to $\sum\limits_{i = 1}^{n-2} i = \frac{(n-2)(n-1)}{2}$.
\end{proof}

Lemma~\ref{lemma:distance_delete_taxon} is important for some of the following proofs using induction.

In most of the following inductive proofs the length of a shortest paths $p$ on trees on $n+1$ taxa is compared with a path on $n$ taxa induced by $p$.
The following lemma bounds the length of these induced paths utilising the rank differences of certain internal nodes between start and end tree of $p$.
Henceforth the notion $\parent(x)$ refers to the internal node that is parent of taxon $x$.

@@

\begin{proof}
    Let $p$ be a shortest path from tree $T$ to $R$ and $p{\big|}_n$ the path resulting from deleting taxon $n+1$ from all trees on $p$.
    This new path $p{\big|}_n$ is then a path from $T$ to $R$.
    Showing that at least $\delta$ $\rnni$ moves are deleted from $p$ to receive $p{\big|}_n$ results in $d(T{\big|}_n,R{\big|}_n) \leq |p{\big|}_n| \leq |p| - \delta = d(T,R) -\delta$ which proves the lemma.
    By the following two observations one can easily see that this claim is true, indeed:
    The rank of an internal node $v$ only changes within an $\rnni$ move if this move is a rank move of $v$ or an $\nni$ move on an edge incident to $v$.
    Furthermore, any $\rnni$ move can change the rank of an internal node by at most one.
    Applying this to $\parent(n+1)$ and knowing that the rank of this node differs by $\delta$ between $T$ and $R$ proves that $d(T{\big|}_n, R{\big|}_n) \leq d(T,R) - \delta$.

@@

Related to Lemma~\ref{lemma:distance_delete_taxon} is Proposition~\ref{proposition:lower_bound_distance} which provides a lower bound for the distance of two trees.
Using the same arguments as in the proof of Lemma~\ref{lemma:distance_delete_taxon}, it easily follows that the rank difference of parents of the same taxon but in two different trees gives such a bound.

@@
